BOLD Parameter Estimation Using Particle Filters

### 2010-07-19 doc/thesis/methods.tex
more to the results

@@

>>>>

\section{Particle Filter Settings}
The run-time for a single voxel depends on the several factors. First, the
overall length of the signal being analyzed. For 1000 measurements it takes
about 6 minutes. On the other hand, in real circumstances the
length is only around 150 measurements and takes around 1 minute. The size of 
local linearization steps can certainly also make a large difference; and 
although it can be tempting to decrease the resolution, going above .001 seconds
per step is not recommended. In most cases millisecond resolution
is fine; however, when generating simulated data I found that every once
in a while this was not enough. This is problematic in the actual particle
filter since, given the large number of simultaneous integrations taking 
place, its probable that a few particles will fail and be unfairly thrown away.
There are two possible outcomes of this. The typical case 
is at some point where are the particles are in a similar location and 
so many or all particles fail together.
The result is then particle deprivation - no particles with non-zero weights remain.
At best this will cause a convergence to a less than optimal posterior 
probability, at worse at will result in a false negative. 
The other possible outcome is that low time constant (or fast moving)
particles get pruned, leaving an excessively smoothed estimate of the output.

Although runge-kutta solvers could be applied, this will still not
guarantee that features aren't missed. Its possible that a
kind of stop-gap measure could be put into place; wherein particles that are
about to be set to NaN are integrated again with finer grained steps. However
many times the non-real results don't occur until several time steps after the 
numbers become unrealistic. So for instance, the time step was too long, allowing 
$f$ to go negative, resulting in extremely large values of $q$. There are many
different ways where this sort of event can occur, and unfortunately sometimes
there is no way to get back to before the state starting going out of control.

Another crucial factor for run time is how long before the first re-sampling 
occurs. Because the prior is represented initially with significantly more
particles, if the model fits very well, or for some reason the effective
number of particles stays high, resampling could take a long time to occur.
For this reason, rather than allowing the particle filter to continue on 
with this large number of particles, after 20 seconds have passed the
algorithm forces resampling. Of course the choice of 20 seconds is somewhat
arbitrary, but at the very least this gives some time for the particle to
be re-weighted before the new distribution is drawn. At the very least the
distribution will be a thinned out version of the prior.

@@

\section{FMRI Configuration}
For the FMRI data discussed in \autoref{sec:RealData}, tests were 
performed on a right handed volunteer using a Gradient EPI sequence
on a 1.5T GE scanner with a 2.1s TR. Slices were 5mm thick and echo
time was (todo).

### 2010-07-19 doc/thesis/results.tex
more to the results

\chapter{Simulation}
Two levels of simulation are discussed in this section. First the simulation of single time series
and second, a slice of simulated FMRI time series. The single time series tests
are necessary to investigate the power of particle filters in 
identifying the BOLD model in a noisy environment. If the particle filter is unable
to get back to the "true" time-series in spite of noise then the particle filter
will not be useful. Additionally, determination of reasonable parameters from the
time-series is also desirable. The reason to then simulate an entire slice is that
the "Possum" FMRI simulation tool models noise more realistically, for instance by
adding spatial correlation. Additionally 
it provides insight into the applicability of particle filters in modeling the BOLD 
signal on a large scale.

@@

>>>>

\chapter{Simulation}
Two levels of simulation are discussed in this section. First the simulation of single time series
and second, a slice of simulated FMRI time series. The single time series tests
are necessary to investigate the power of particle filters in 
identifying the BOLD model in a noisy environment. If the particle filter is unable
to get back to the "true" time-series in spite of noise then the particle filter
will not be useful. Additionally, determination of reasonable parameters from the
time-series is also desirable. The reason to then simulate an entire slice is that
the "Possum" FMRI simulation tool models noise more realistically, for instance by
adding spatial correlation. Additionally 
it provides insight into the applicability of particle filters in modeling the BOLD 
signal on a large scale. Both sorts of tests are usefull in determining the
consistency of the particle filter. For the single voxel tests, each test will
actually be performed eleven different times, each with a new noise realization.
If runs fail to have consistent estimates of the underlying BOLD signal, then
this would indicate a large variance error in the model. Similarly, in the
Possum simulation there are actually only four discrete parameter sets driving the time-series.
Nevertheless each voxel will have a different tissue composition and a different
noise realization. Therefore the Possum simulation should also provide a good
test of the consistency of the BOLD model.

@@

There were two primary purposes of these tests. First given the nature of monte-carlo
techniques it is important to ensure consistency of results. Although the parameter
sets were somewhat inconsistent, the quality of the output was actually quite consistent
across eleven runs. The second purpose was of course to determine how the particle 
filter responded to different signal to noise ratios. Although the direct $\sqrt{MSR}$ is not
a good measure of quality, the normalized $\sqrt{MSR}$ is. Drawing a line at a particular
threshold may be difficult, however from the tests it seems that normalized $\sqrt{MSR}$
below $1$ may be a good start.

### 2010-07-19 doc/thesis/resultsB.tex
more to the results

\chapter{Real Data}
Finally, we also performed inference on a real FMRI scan. The scanner we used...
%... more specifics...

Before performing tests on a full image, I tested the results of the particle filter
on regions deemed active and non-active by statistical parametric mapping. 
This served the purpose of tuning the priors to reasonable levels etc.
After that, the particle filter was applied to every voxel in an FMRI 
image individualy, and the results were put into a parameter map 
image. Additionally, the resulting MSE and square-root MSE were placed
into images for further analysis.

\section{Single-Voxel Analysis}
This section discusses the results when the particle filter was
applied on a single voxel. The parameters are the same as
those used later for entire image analysis; however, the results
are more in-depth.

The run-time for a single voxel depends on the several factors. First, the
overall length of the signal being analyzed. For 1000 measurements it takes
about 10 to 15 minutes. On the other hand, in real circumstances the
length is only around 150 measurements. The number of  integration
can certainly make a large difference, however dropping below 1000 (.001 seconds)
is definitely not recommended.

For a period I considered 1000 to be a
fine number; however when generating simulated data I found that every once
in a while 1000 was not enough. This is problematic in the actual particle
filter since, given the large number of simultaneous integrations taking 
place, its likely that a few particles will fail and be weighted at 0 because
of this problem. Additionally, because the typical case where a failure would
occur is at fast moving times/parameters the particles all tend to fail together.
The result is particle deprivation - no particles with non-zero weights remain.
The other possible outcome is that low time constant particles get pruned resulting
in excessively smoothed estimates for the time series'. Its possible that a
kind of stop-gap measure could be put into place; wherein particles that are
about to be set to NaN are integrated again with finer grained steps. However
many times the non-real results don't occur until several time steps after the 
numbers get strange. So for instance, the timestep was too long, allowing 
$f$ to go negative, resulting in extremely large values of $q$. There are many
different ways where this sort of event can occur, and unfortunately sometimes
there is no way to get back to before the state starting going out of control.

Another crucial factor for run time is how long before the first re-sampling 
occurs. Because the prior is represented initially with significantly more
particles, if the model fits very well, or for some reason the effective
number of particles stays high, resampling could take a long time to occur.
When this happens the particles filter can take a factor of 10 longer to run.
However, if the particle count isn't initially set high, there is a much larger
chance of particle deprivation occurring. Since there is no real way to know
how long it will take to resmaple the first time, there is little the 
algorithm can do to fix this (except perhaps forcing resampling after
some period of time). On the other end of the spectrum, if the time
series doesn't match the model at all, particle deprivation will occur extremely
quickly.  The upshot of this is that the particle filter is able to 
identify these sections very quickly, and thus not waste much time there.
The difficulty though, is the regions in between the perfect fit and the
awful fit. If particle deprivation does occur, did it occur randomly to an
activated region or did it occur inevitably because the region doesn't fit.
The reason for having so many initial particles is to give density to the
distribution to make false negatives more rare. Perhaps the correct method
is to still set the initial particles very high, but if for a long time no
resampling occurs, halve the standard deviation of the measurement error.
Thus, if the weighting function is not discriminating enough, force it 
to be more picky about the results. Of course this could lead to particle
deprivation as well if the standard deviation is brought down too quickly.
Of course, if a fat-tailed distribution is used for the weighting function,
or the standard deviation of a Gaussian weighting function is sufficiently large,
the particle filter will simply converge to meaningless values. The question
of whether

\begin{figure}
\includegraphics[clip=true,trim=10cm 4cm 10cm 4cm, width=16cm]{images/inactive_illogical}
\caption{Particle Filter converging to values that make little sense,
because the voxel did not correlate with the input in any known way}
\label{fig:badfit}
\end{figure}

\begin{figure}
\includegraphics[clip=true,trim=10cm 4cm 10cm 4cm, width=16cm]{images/active_difficult}
\caption{Example of a timeseries that laplace performs better on than
gaussian. The large spike at the beginning can quickly eliminate all particles} 
\label{fig:highhard}

\end{figure}

For this reason, it is important to threshold based on the MSE, and the
impulse response. Naturally the MSE is intended to catch very voxels with a
very poor fit,
but like \autoref{fig:badfit}, the algorithm may have found a line
down the middle that gives a decent mean squared error. In this case
the excessive smoothing and low actual activation level will easily
be distinguished by finding the peak response to single short pulse.

The choice of a prior, as discussed previously, is extremely important. While a
prior may have the potential to give good results, being a monte-carlo algorithm
there is the possibility for inconsistencies. Thus, increasing the variance
of the time-constants may allow additional flexibility, it will also cause
additional model variance. Case in point, the exact same algorithm run twice
with standard deviations of $.35, .35, .35$ for the time constants resulted in two
very different fits, see \autoref{fig:param1_var}.

\begin{figure}
\subfigure[]{\includegraphics{images/badfit_param1}}
\subfigure[]{\includegraphics{images/goodfit_param1}}
\caption{The same priors gave rise to both fits. Params: ($\tau_0 \alpha E_0 V_0 \tau_f \tau_s \epsilon$)
for the good fit: $(1.40 .32 0.34 0.04  0.96 2.76)$, and for the poor fit:
$(2.28 0.32 0.33 0.017 0.67 2.88 1.36)$.}
\label{fig:badfit_param1}
\end{figure}

For this reason, I actually lowered the standard deviattions of the time
constants to prevent over-smoothing. This resulted in the more consistent, but
slightly worse results seen in \autoref{fig:param2_var}

\begin{figure}
\subfigure[]{\includegraphics{images/param2a}}
\subfigure[]{\includegraphics{images/param2b}}
\caption{A poor fit, using the same parameters as }
\label{fig:param2_var}
\end{figure}

\section{Particle Collapse Recovery}
This heuristic was initially intended to deal with cases where the particle
filter prematurely converged; and to reverse this effect. In reality, however
it did not perform well, and so it was not used in full-brain analysis. Ultimately
it did not prevent collapse of the distribution, and it often caused dramatic
increased in the variance of parameter estimates. It also pronged the analysis
time of inactive time series. Essentially the effect was exactly opposite of what
one would hope.

\section{Single Time-Series Simulation}

Graphs:

For simulated data, single timeseries:

For \{delta, DC/Spline\}, \{exponential, gaussian, cauchy\}, \{biased, unbiased initial\},
\{100, 500, 1000\} particles
\begin{enumerate}
\item Ground truth vs. Estimated signal during particle filter run
\item Ground truth vs. Estimated signal with final parameter set
\item True Parameters vs. Final Parameter Sets
\item Variance of final parameters when faced with same ground truth, different noise
\item MSE of (a new timeseries based on X(t) vs. ground truth) for all t
\item Estimator Variance based on different noise runs
\item Final Particle Distribution
\end{enumerate}

For Simulated Data, Full Volume:

%note to self, epsilon should probably be uniform between 0 and something
\section{Simulated Volume}

\chapter{Real Data}
\begin{figure}
\subfigure[]{\label{fig:hm_spm} \includegraphics[scale=.66]{images/spm_hm}}
\subfigure[]{\label{fig:hm_canon_spm_x} \includegraphics[scale=.85]{images/spm_hm_x}}
\subfigure[]{\label{fig:hm_canon_spm_y} \includegraphics[scale=.85]{images/spm_hm_y}}
\subfigure[]{\label{fig:hm_canon_spm_z} \includegraphics[scale=.85]{images/spm_hm_z}}
\subfigure{\label{fig:scale_spm} \includegraphics[scale=.3]{images/scale1}}
\caption{Sagittal, coronal and axial slices of SPM results (\autoref{fig:hm_canon_spm_x} \autoref{fig:hm_canon_spm_y} 
         \autoref{fig:hm_canon_spm_x}), as well as a series of axial slices, \autoref{fig:hm_spm}. Units
         of activation are in Student's T-scores. Higher indicates higher assurance that the signal cannot
         have occurred through noise alone.}
\label{fig:hm_canon_spm}
\end{figure}

@@

image comparing epsilon-map with GLM activation map

>>>>

%\chapter{Real Data}
%Finally, we also performed inference on a real FMRI scan. The scanner we used...
%... more specifics...

% TODO include single?
%Before performing tests on a full image, I the particle filter
%on regions deemed active and inactive by statistical parametric mapping
%(SPM). This served the purpose of adjusting the priors as well as the 
%preprocessing based on real world signals. This was actually done before
%the simulations, and then results were carried back the simulations 
%to check consistency.
%After work adjusting parameters, most importantly the weighting function and the 
%priors, particle filter was applied to every voxel in an FMRI image.
%The results of this large scale analysis was a parameter map which was
%then used to calculate normalized square-root MSE image. 
%
%\section{Single-Voxel Analysis}
%The choice of a prior, as discussed previously, is extremely important. While a
%prior may have the potential to give good results, being a monte-carlo algorithm
%there is the possibility for inconsistencies. Thus, increasing the variance
%of the time-constants may allow additional flexibility, it will also cause
%additional model variance. Before running on a full volume I adjusted the 
%prior to ensure that the same input would give the same output 100 times in a 
%row. While this may seem like a given, with a random drawing of the prior,
%this can be difficult. Case in point, the exact same algorithm run twice
%with the time constants all having standard deviations of $.35$ resulted in two
%very different fits, shown in \autoref{fig:badfit_param1}.
%
%\begin{figure}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/badfit_param1}}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/goodfit_param1}}
%\caption{The same priors gave rise to both fits.}
%\label{fig:badfit_param1}
%\end{figure}
%
%For this reason, I actually lowered the standard deviations of the time
%constants to prevent over-smoothing. This resulted in more consistent,
%though potentially slightly worse fits, two examples of which are 
%shown in \autoref{fig:param2_var}. 
%
%\begin{figure}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/param2a}}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/param2b}}
%\caption{A poor fit, using the same parameters as }
%\label{fig:param2_var}
%\end{figure}
%
%todo: stats of the 100 fits?
%
%\section{Single Time-Series Simulation}
%
\chapter{Real Data}
\label{sec:RealData}
Modeling the BOLD response is of course not of much use if it is only done in a 
single voxel. Although this algorithm will hopefully lead to more novel methods 
of analysis, the standard use for modeling the BOLD signal is to locate "activation".
By activation I really mean areas where the input correlates with the BOLD response, 
thus indicating that the stimuli is driving some sort of neural activity. Of course 
much more is going on behind the scenes, brain regions are stimulating other brain
regions, but before that can be analyzed we must validate the ability of the particle
filter to be able to accurately find and model activation. Once areas where the BOLD 
model may be accurately estimated are found, integrating the model will allow for accurate
simulation between measurements, which then may be used for more advanced analysis.
Again though, it all begins with localizing the first activation regions in the
chain. Therefore here I compare the output of the particle filter based modeling
solution to the conventional SPM method.

In reality, the SPM8 method is a different animal from the algorithm described here.
First SPM preprocesses the image by spatially smoothing the FMRI image (here with 
an $8mm x 8mm x 8mm$ Gaussian kernel), whereas
this is not done in the particle filter algorithm. Additionally, a spline
is used to de-trend, rather than SPM8's high pass filter with a cut
off based on a globally estimated autocorrelation. Thus the preprocessing pipelines 
are different, and the output of SPM8 is single t-statistic, whereas the output 
of the particle filter is a posterior probability distribution of the parameters
at every single voxel. To validate the quality of the particle filter results though
it is necessary to compare the location of
"activated" voxels, and the goodness of fit provided by each method.

The results from SPM8 are shown in \autoref{fig:hm_canon_spm}, and the results from 
the particle filter are shown in \autoref{hm_canon_pfilter}.

\begin{figure}
\subfigure[]{\label{fig:hm_spm} \includegraphics[scale=.66]{images/spm_hm}}
\subfigure[]{\label{fig:hm_canon_spm_x} \includegraphics[scale=.85]{images/spm_hm_x}}
\subfigure[]{\label{fig:hm_canon_spm_y} \includegraphics[scale=.85]{images/spm_hm_y}}
\subfigure[]{\label{fig:hm_canon_spm_z} \includegraphics[scale=.85]{images/spm_hm_z}}
\subfigure{\label{fig:scale_spm} \includegraphics[scale=.3]{images/scale1}}
\caption{Sagittal, coronal and axial slices of SPM results (\autoref{fig:hm_canon_spm_x} \autoref{fig:hm_canon_spm_y} 
         \autoref{fig:hm_canon_spm_x}), as well as a series of axial slices, \autoref{fig:hm_spm}. Units
         of activation are in Student's T-scores. Higher indicates higher assurance that the signal cannot
         have occurred through noise alone.}
\label{fig:hm_canon_spm}
\end{figure}

@@
